---
title: "34474056_A2"
author: "Chan Long Hei"
date: "2026-09-6"
output: html_document
---

# FIT5145 Assignment 2

## Loading Libraries and Datasets

```{r}
library(tidyverse)
library(stringr)
library(scales)

# Read files
wastes <- read_csv("Wastes.csv")
year_state <- read_csv("Year_State_ID.csv")
```

## Initial Inspection of Datasets

```{r}
# Roughly look at the datasets
glimpse(wastes)
glimpse(year_state)
```

```{r}
# Looking at summary to see if anything is out of line
summary(wastes)
summary(year_state)
```

```{r}
# Looking for NA entries for each column in wastes
colSums(is.na(wastes))

```

```{r}
# Looking for NA entries for each column in year_state
colSums(is.na(year_state))
```

## **Q1**

**How many unique “Category” values are there in the data file (6 marks)?**

### 1a. Inspecting raw data for Category values

```{r}
# Inspect all Category values and their frequencies
wastes %>%
  count(Category, sort = TRUE)
```

### 1b. Checking other Types to reassign some suspicious Category values

```{r}
# Extracting suspicious Category
suspicious_categories <- c(
  "organics",
  "Plastic",
  "Building & demolition materials",
  "HW",
  "Toxic wastes",
  "Aluminium",
  "Metals & Organics",
  "Recycling",
  "Space debris"
)

# Inspecting suspicious Category values
wastes %>%
  filter(
    Category %in% suspicious_categories |
      is.na(Category)
  ) %>%
  select(
    Case_ID,
    Category,
    Type,
    Stream,
    Fate,
    Tonnes
  )
```

```{r}
# Find the Types associated with the suspicious Category values
suspicious_types <- wastes %>%
  filter(Category %in% suspicious_categories) %>%
  distinct(Category, Type) %>%
  arrange(Category, Type)

suspicious_types
```

```{r}
# Verifying the common Category associated with these types
wastes %>%
  semi_join(
    suspicious_types,
    by = "Type"
  ) %>%
  count(Type, Category) %>%
  arrange(Type, desc(n))
```

```{r}
# Find Types where Category is missing
missing_category_types <- wastes %>%
  filter(is.na(Category)) %>%
  distinct(Type)

missing_category_types
```

```{r}
# Check where those Types normally belong
wastes %>%
  semi_join(
    missing_category_types,
    by = "Type"
  ) %>%
  filter(!is.na(Category)) %>%
  count(Type, Category, sort = TRUE)
```

### 1c. Cleaning the Category values

```{r}
wastes_clean <- wastes %>%
  mutate(
    Category = case_when(
      
      # Naming inconsistencies
      Category == "organics" ~ "Organics",
      Category == "Plastic" ~ "Plastics",
      Category == "Building & demolition materials" ~
        "Building and demolition materials",
      
      # Wrongly classified values verified using Type
      Category == "Aluminium" ~ "Metals",
      Category == "Metals & Organics" ~ "Organics",
      Category == "Recycling" ~ "Plastics",
      Category == "HW" ~ "Hazardous wastes",
      Category == "Toxic wastes" ~ "Hazardous wastes",
      
      # Missing categories that can be inferred from their Type
      is.na(Category) &
        Type == "Other soil/sludges (other N)" ~
        "Hazardous wastes",
      
      is.na(Category) &
        Type == "Paints, resins, inks, organic sludges (F)" ~
        "Hazardous wastes",
      
      # Keep all otherwise valid values unchanged
      TRUE ~ Category
    )
  )

# Removing the entry where Category is Space Debris
wastes_clean <- wastes_clean %>%
  filter(
    !(Category == "Space debris" &
        is.na(Case_ID) &
        is.na(Year_State_ID))
  )
```

### 1d. Calculate the number of unique Categories

```{r}
# Showing all unique Categories
wastes_clean %>%
  distinct(Category) %>%
  arrange(Category)
```

```{r}
# Showing number of unique Categories
wastes_clean %>%
  summarise(
    unique_categories =
      n_distinct(Category, na.rm = TRUE)
  )
```

### Q1 Answer and Justification

The initial inspection shows that there are 21 distinct values in the raw Category column including NA. However, several values represented inconsistent naming or apparent data-entry errors rather than genuinely different waste categories. For example, organics and Organics, Plastic and Plastics, and the two forms of building and demolition materials referred to the same categories. Other suspicious values were checked against their corresponding Type. I have decided to impute them from other types instead of removing them from the dataset. Aluminium was reassigned to Metals, which is supported both by other observations and by the specification, which describes Aluminium as a Type within the Metals category. Similarly, Recycling was reassigned to Plastics because its Type was PVC and the specification defines recycling as a Fate rather than a Category. Missing Category values were assigned to Hazardous wastes where the same Types were consistently classified as hazardous waste elsewhere in the data. One incomplete Space debris record was excluded because it has more than 2 missing variables, with no case ID, year/state ID, stream, fate, tonnage or description and therefore could not be meaningfully used in the analysis. After these corrections, the dataset contains **11 unique waste Categories**.

## Q2

**How many negative feedback comments are in the 'Description' column with an environmental impact score of 2 or 3 (4 marks)?**

From initial inspection of the raw dataset and data cleaning in Q1, it was found that there is 1 NA entry in the Description column in the raw dataset, which is associated with the entry with Category "Space Debris". It was already removed back in Q1, therefore we can conclude that there is no more NA entry in the Description column.

### 2a. Inspection of Description structure

```{r}
# Extracting 10 rows to inspect the Description structure
wastes_clean %>%
  select(Description) %>%
  slice_head(n = 10)
```

### 2b. Extracting Environmental Impact Score

```{r}
q2_data <- wastes_clean %>%
  mutate(
    Impact_Score = as.numeric(
      str_extract(
        Description,
        "(?<=Environmental Impact Score: )[0-9]+"
      )
    )
  )

q2_data
```

### 2c. Extracting negative feedback comments

```{r}
# Showing number of rows in each of the impact score of 2,3
q2_data %>%
  filter(Impact_Score %in% c(2, 3)) %>%
  count(Impact_Score)
```

```{r}
# Roughly looking at the comment description for impact score of 2,3
q2_data %>%
  filter(Impact_Score %in% c(2, 3)) %>%
  select(Description) %>%
  slice_head(n = 20)
```

```{r}
# Calculating total number of negative feedback comments
q2_data %>%
  filter(Impact_Score %in% c(2, 3)) %>%
  summarise(
    negative_feedback_comments = n()
  )
```

### Q2 Answer and Justification

Since the only missing Description record had already been removed during the earlier cleaning stage in Q1, no additional missing-value treatment was required. The environmental impact score was extracted from the text in Description, then records with scores of 2 or 3 were filtered and counted. The description of comments with impact score of 2,3 were also inspected. The rough inspection found that all the filtered descriptions are negative comments, which suggests that all comments with impact score of 2,3 are negative comments as it is a low impact score which negatively affects the environment as stated in the specifications. Thus, there are a total of **6312 negative feedback comments** in the Description column with an environmental impact score of 2 or 3.

## Q3

**For each Category value, write code to calculate the fractions of waste tonnes of different waste sources, and then draw a chart to visualise the fraction numbers specific to the Category value (4 marks).**

### 3a. Data inspection and cleaning

From the initial inspection of the dataset, it was found that there were 15 missing Tonnes values. It was also found that there are negative Tonnes values, which should be further investigated.

```{r}
# Calculating total missing Tonnes values and their contribution to the entire dataset
wastes_clean %>%
  summarise(
    missing_tonnes = sum(is.na(Tonnes)),
    missing_pct = mean(is.na(Tonnes)) * 100
  )
```

```{r}
# Inspecting negative Tonnes values
wastes_clean %>%
  filter(Tonnes < 0) %>%
  select(
    Case_ID,
    Category,
    Type,
    Stream,
    Fate,
    Tonnes,
    Description
  )
```

From the initial inspection of the dataset, it was found that there was 1 missing Stream value, which was removed back in Q1. Therefore, no further NA checking has to be done.

```{r}
# Inspecting the Stream variable
wastes_clean %>%
  count(Stream)
```

```{r}
# Removing the entries with missing Tonnes, Replacing the odd negative Tonnes value and cleaning up the inconsistent Stream entries
wastes_clean <- wastes_clean %>%
  mutate(
    Stream = case_when(
      Stream == "Municipal solid waste" ~ "MSW",
      Stream == "Construction and demolition waste" ~ "C&D",
      TRUE ~ Stream
    ),
    Tonnes = case_when(
      Case_ID == 34666 ~ 141663.39,
      Tonnes < 0 & abs(Tonnes) < 0.001 ~ 0,
      TRUE ~ Tonnes
    )
  ) %>%
  filter(!is.na(Tonnes))
```

### 3b. Calculating Waste Tonnes Fractions

```{r}
# Calculate total tonnes and fractions by Category and Stream
q3_summary <- wastes_clean %>%
  group_by(Category, Stream) %>%
  summarise(
    Total_Tonnes = sum(Tonnes),
    .groups = "drop"
  ) %>%
  group_by(Category) %>%
  mutate(
    Fraction = Total_Tonnes / sum(Total_Tonnes)
  ) %>%
  ungroup() %>%
  mutate(
    Category = factor(
      Category,
      levels = sort(unique(Category))
    )
  )

# Display the calculated fractions
q3_summary
```

```{r}
# Check fractions sum to 1 within every Category
q3_summary %>%
  group_by(Category) %>%
  summarise(
    Total_Fraction = sum(Fraction)
  )
```

### 3c. Visualising the fractions

```{r}
# Visualise the fractions
ggplot(
  q3_summary,
  aes(
    x = Category,
    y = Fraction,
    fill = Stream
  )
) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(
    labels = scales::percent_format(),
    limits = c(0, 1)
  ) +
  labs(
    title = "Fraction of Waste Tonnes by Source for Each Category",
    x = "Waste Category",
    y = "Fraction of Total Waste Tonnes",
    fill = "Waste Source"
  ) +
  theme_minimal()
```

### Q3 Answer and Justification

Before calculating the fractions, the Tonnes variable was checked for missing values. Fifteen observations had missing tonnage values, representing approximately 0.04% of the dataset. Since only a very small fraction of cases was affected, these observations were removed rather than imputed.

Inspection of negative Tonnes values identified one odd case: Case 34666 contained a value of -20,000,000 tonnes, while its Description stated that 141,663.39 tonnes were processed. Since the description provided a clear corresponding quantity, the suspicious value was corrected to 141,663.39 rather than removing the observation. The remaining negative values were extremely small and close to 0, with their descriptions reporting -0.00 tonnes. These were treated as numerical precision errors and replaced with 0 because a negative waste quantity is not meaningful.

The Stream variable was standardised before analysis because the dataset contained duplicate labels for the same waste source, such as Municipal solid waste and MSW, and Construction and demolition waste and C&D. Since the specification defines only three stream categories, MSW, C&I, and C&D, these duplicate labels were recoded into the standard forms. This prevents the same waste source from being treated as separate groups when calculating fractions.

Total tonnes were then calculated for each combination of Category and Stream, and each stream total was divided by the total tonnes within its Category to obtain the fraction. A 100% stacked bar chart was selected because both Category and Stream are categorical variables and the values being visualised are proportions that sum to 100% within each Category. This allows the composition of waste sources to be compared directly across waste categories.

## Q4

**Please add the 'Year' and 'State' values from Year_State_ID.csv to Wastes.csv, compute the total waste tonnes for each year and state, and store the result in a dataset named 'temp'. Then, use a single R function to compute and display the statistical information (i.e., Min, Max, and Mean) of the total waste tonnes for each state in ‘temp’ (Note: You may use multiple functions/commands to prepare the pre-processed data table, but when you compute and display the statistical information, you need to use a single R function.) (4 marks).**

### 4a. Inspecting and cleaning year_state dataset

From the initial inspection of the dataset, it was found that there were no NA values for all columns in year_state. Therefore, no additional checks for NAs were needed.

```{r}
# Inspect all State values and their frequencies
year_state %>%
  count(State, sort = TRUE)
```

No duplicate states with inconsistent format are present. However, Australia is not a state but a country. The specification mentions that the State column is the state or territory in which the waste was generated.

```{r}
# Looking at Australia related data
aus_data <- year_state %>%
  filter(State == "Australia")

aus_data
```

The entries related to "Australia" as the state was also lacking, contributing only to 2 distinct year ranges.

```{r}
# Determining if there are duplicate ID values
n_distinct(year_state$ID)
nrow(year_state)
```

All year_state IDs are distinct.

```{r}
# Inspect all Year values and their frequencies
year_state %>%
  count(Year, sort = TRUE)
```

All Year values are formatted consistently. However, the specification mentions that Year is the financial year for each year between 2006-2007 and 2020-2021. It was found that there are years exceeding the given range.

```{r}
# Check if all Year_State_ID from wastes match to an existing ID in year_state
wastes_clean %>%
  # anti_join is used to identify mismatches found for the foreign key (wastes$Year_State_ID)
  anti_join(
    year_state,
    by = c("Year_State_ID" = "ID")
  )
```

All Year_State_ID in wastes has an existing match to an ID in year_state.

**Data Cleaning**

```{r}
# Removing invalid Year ranges from year_state
year_state_clean <- year_state %>%
  separate(Year, into = c("start_year", "end_year"),
           sep = "-", remove = FALSE) %>%
  mutate(
    start_year = as.numeric(start_year),
    end_year = as.numeric(end_year)
  ) %>%
  filter(end_year - start_year == 1, 
         start_year >= 2006)

# Verifying cleaned Year ranges
year_state_clean %>%
  count(Year, sort = TRUE)
```

```{r}
# Removing entries with State as "Australia"
year_state_clean <- year_state_clean %>%
  filter(!State == "Australia")

year_state_clean %>%
  count(State, sort = TRUE)
```

### 4b. Joining the datasets

```{r}
# Joining the datasets
wastes_joined <- wastes_clean %>%
  left_join(
    year_state_clean,
    by = c("Year_State_ID" = "ID")
  )

# Removing any NA rows for wastes_joined
wastes_joined <- na.omit(wastes_joined)
```

```{r}
# Roughly look at the joined dataset
glimpse(wastes_joined)
summary(wastes_joined)
```

### 4c. Creating temp and displaying the statistical information

Tonnes values were already cleaned back in Q3, therefore no additional cleaning is needed before computing total waste tonnes.

```{r}
# Computing total waste tonnes for each year and state
temp <- wastes_joined %>%
  group_by(Year, State) %>%
  summarise(
    Total_Tonnes = sum(Tonnes),
    .groups = ("drop")
  )

temp
```

```{r}
# Reshaping temp so that each State is a column
temp <- temp %>%
  select(Year, State, Total_Tonnes) %>%
  pivot_wider(
    names_from = State,
    values_from = Total_Tonnes
  ) %>%
  select(-Year)

temp
```

```{r}
# Showing the statistical information of temp with a single function
summary(temp)
```

### Q4 Answer and Justification

The Year_State_ID dataset was first inspected to determine if any cleaning is needed. The national-level observations labelled “Australia” were excluded because the analysis requires comparisons across individual states and territories. Additionally, year ranges spanning more than one financial year and years out of range were also removed as inconsistent temporal observations as the specifications state so.

Wastes.csv and Year_State_ID.csv were then combined using a left join, matching Year_State_ID in the waste data with ID in the year-state data. A left join was used because Wastes.csv is the main dataset and the objective is to add the corresponding Year and State values from Year_State_ID.csv while retaining all waste observations. Year_State_ID in Wastes.csv acts as the foreign key, while ID in Year_State_ID.csv is the primary key. The joined data were grouped by Year and State to calculate total waste tonnes and stored as temp. Finally, summary() was used as the single R function to display the minimum, mean, and maximum total waste tonnes for each state.

## Q5

**A manager wants to see a chart showing a yearly trend of total waste tonnes of food organics for each state. To draw the chart, please convert the Year-Year formats of all Year values into Year formats (4 marks) (Note: Use the dataset with the converted year format for all subsequent questions.)**

Back in Q3 and Q4, data inspection and cleaning was already done for both the Tonnes column and the Year column. Therefore, only the Type column is left unchecked.

### 5a. Converting Year Format

```{r}
# Converting Year column format (Year-Year to Year)
wastes_joined <- wastes_joined %>%
  mutate(Year = as.numeric(substr(Year, 1, 4)))

# Inspecting Year column after conversion
wastes_joined %>%
  count(Year, sort = FALSE)
```

### 5b. Inspecting and cleaning Type values

```{r}
# Look at distinct Type values
wastes_joined %>%
  count(Type, sort = FALSE)
```

After inspecting the Type values, it was found that all Types are distinct and there are no suspicious Types which should be represented differently. It was also found back in initial inspection of dataset that there is no NA values under Type. Therefore, no cleaning is needed.

### 5c. Filtering for Food Organics

```{r}
# Filter out total tonnes of each state and year for food organics
q5_data <- wastes_joined %>%
  filter(Type == "Food organics") %>%
  group_by(Year, State) %>%
  summarise(
    Total_Tonnes = sum(Tonnes, na.rm = TRUE),
    .groups = "drop"
  )

q5_data
```

### 5d. Plotting yearly trend of food organics

```{r}
# Plotting line graph to show trends of food organics wastes of each state over the years
ggplot(q5_data,
       aes(x = Year,
           y = Total_Tonnes,
           colour = State,
           group = State)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Yearly Trend of Total Food Organics Waste by State",
    x = "Year",
    y = "Total Waste Tonnes",
    colour = "State"
  ) +
  theme_minimal()
```

### Q5 Answer and Justification

At the start, no cleaning or inspection was needed for Tonnes and Year values as it was already done in previous questions. The financial-year values were converted from Year-Year format to a single starting year to show a yearly trend and ensure that the converted dataset can be used for subsequent questions. The Type column was inspected before filtering to ensure that "Food organics" was consistently recorded throughout the dataset. The data were then filtered to Type == "Food organics", grouped by Year and State, and total Tonnes were summed for each group. A line chart was used because the objective is to compare how food-organics waste changes over time across states.

## Q6

**Write code to display the most recycled waste Type and the most energy-recovered waste Type with the corresponding year. Write code to display the most increased waste Type over years (5 marks).**

For this question, the only column left unchecked and not cleaned is the Fate variable.

### 6a. Inspecting and cleaning Fate column

```{r}
wastes_joined %>%
  count(Fate, sort = TRUE)
```

Inspection of the Fate column shows that all Fates are distinct and there are no suspicious Fates which should be represented differently. Therefore, no cleaning is needed.

### 6b. Most recycled waste Type

```{r}
# Calculating the most recycled waste Type and its corresponding Year
most_recycled <- wastes_joined %>%
  filter(Fate == "Recycling") %>%
  group_by(Type, Year) %>%
  summarise(
    Total_Tonnes = sum(Tonnes, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  slice_max(
    order_by = Total_Tonnes,
    n = 1,
    with_ties = FALSE
  )

most_recycled
```

### 6c. Most energy-recovered waste Type

```{r}
# Calculating the most energy-recovered waste Type and its corresponding Year
most_energy_recovered <- wastes_joined %>%
  filter(Fate == "Energy recovery") %>%
  group_by(Type, Year) %>%
  summarise(
    Total_Tonnes = sum(Tonnes, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  slice_max(
    order_by = Total_Tonnes,
    n = 1,
    with_ties = FALSE
  )

most_energy_recovered
```

### 6d. Most increased waste Type over years

Increase =Total tonnes at end of period − Total tonnes at start of period

= Total tonnes at 2020 - Total tonnes at 2006

Before proceeding with the calculations, it is important to first identify if all the existing Types has data in both the start and end period. The analysis would be incorrect if Types with missing data at one or both ends are included, as they would provide an incorrect waste growth throughout the years, such as missing start year data having 0 as its Tonne value, thus providing a substantially large growth if used to calculate the increase over years.

```{r}
# Determining if every Type appears at both ends of the years
type_check <- wastes_joined %>%
  filter(Year %in% c(2006, 2020)) %>%
  distinct(Type, Year) %>%
  count(Type, name = "Years_Available")

type_check %>%
  count(Years_Available)
```

The Years_Available shows how many years of data are available for each Type. As observed, there are 5 Types with data available in only one end of the years, thus must be excluded from the final calculations to avoid computing false and misleading waste increase.

```{r}
# Filtering out valid types for checking
valid_types <- type_check %>%
  filter(Years_Available == 2) %>%
  pull(Type)

valid_types
```

```{r}
# Calculating Increase in waste of all Types
type_increase <- wastes_joined %>%
  filter(
    Type %in% valid_types,
    Year %in% c(2006, 2020)
  ) %>%
  group_by(Type, Year) %>%
  summarise(
    Total_Tonnes = sum(Tonnes, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(Type) %>%
  summarise(
    Start_Tonnes = Total_Tonnes[Year == 2006],
    End_Tonnes = Total_Tonnes[Year == 2020],
    Increase = End_Tonnes - Start_Tonnes,
    .groups = "drop"
  )

type_increase
```

```{r}
# Computing the Type with most increased waste over the years
most_increased <- type_increase %>%
  slice_max(Increase, n = 1, with_ties = FALSE)

most_increased
```

### Q6 Answer and Justification

The Fate column was first inspected to confirm the exact categories used for recycling and energy recovery and to check for missing or inconsistent values. Total tonnes were then aggregated by Type and Year because multiple records can contribute to the same waste Type within a year. The most recycled and most energy-recovered waste Types were identified by selecting the Type–Year combination with the highest total tonnes for the corresponding Fate. For the “most increased” waste Type, the analysis compared total tonnes at the beginning and end of the cleaned study period rather than using year-to-year changes. Before calculating this change, an endpoint availability check was performed. Of the waste Types, 46 appeared in both the start and end years, while 5 appeared at only one endpoint. The incomplete Types were excluded to avoid computing incorrect waste growth. The Type with the largest positive difference between end-period and start-period total tonnes was therefore identified as the most increased waste Type over the full period.

The most **recycled** waste Type is **Bricks, concrete and pavers**, in year **2020** with a total Tonnes of **10478681**.

The most **energy-recovered** waste Type is **Food organics**, in year **2015** with a total Tonnes of **1140998**.

The **most increased** waste Type over years is **Bricks, concrete and pavers**, with a total increase of **8345529** tonnes over the observed years.

## Q7

**Please investigate the factors influencing environmental impact scores. (Note: You may use variables from both Wastes.csv and Year_State_ID.csv for your analysis.) Marks will be awarded based on the depth and breadth of your investigation, including analyses and discussions, as well as the strength of your reasoning in drawing insights and conclusions (5 marks).**

### 7a. Extract and Inspect Impact Score

```{r}
# Extract Impact Score from Description
wastes_joined <- wastes_joined %>%
  mutate(
    Impact_Score = as.numeric(
      str_extract(
        Description,
        "(?<=Environmental Impact Score: )[0-9]+"
      )
    )
  )

# Inspect count of each Impact Score 
wastes_joined %>%
  count(Impact_Score, sort = FALSE)

# Determine if there are any rows with NA Impact Score
wastes_joined %>%
  summarise(
    NA_Impact_Score = sum(is.na(Impact_Score)),
  )

```

The extracted environmental impact scores contained no missing values and all observed scores fell within the valid 1–10 range. However, the distribution was highly concentrated at a few values rather than being evenly spread across the scale. Score 5 was the most common with 16,100 observations, followed by score 6 with 8,798, score 3 with 6,304 and score 8 with 5,476. Scores 1, 2 and 4 occurred only rarely, while scores 7, 9 and 10 were not observed. This concentration should be considered when interpreting subsequent comparisons, as changes in group averages may be driven by different proportions of these dominant score levels.

### 7b. Investigating Factors Affecting Impact Score

For this analysis, the chosen factors to investigate are as such: Fate, Stream, Category, Tonnes, Core_Non-score, Year, State, Economic_Growth. The factors were chosen because they represent different dimensions that may be associated with environmental impact scores. Fate captures how waste is ultimately managed, while Stream and Category describe the source and broad type of waste. Tonnes represents the scale of the waste process. Core_Non-core distinguishes core waste from mining, agriculture and such non-core waste, which provides another meaningful structural classification. State captures possible geographic differences in waste-management practices, while Year captures possible impact score changes through time. Economic_Growth allows broader economic conditions to be considered. Together, these variables provide a broad investigation using information from both datasets.

#### 7b(i) Check all variables used

```{r}
# Check the variables used for na entries
wastes_joined %>%
  summarise(
    Missing_Impact_Score = sum(is.na(Impact_Score)),
    Missing_Tonnes = sum(is.na(Tonnes)),
    Missing_Fate = sum(is.na(Fate)),
    Missing_Stream = sum(is.na(Stream)),
    Missing_Category = sum(is.na(Category)),
    Missing_State = sum(is.na(State)),
    Missing_Year = sum(is.na(Year)),
    Missing_Economic_Growth = sum(is.na(Economic_Growth)),
    Min_Impact_Score = min(Impact_Score, na.rm = TRUE),
    Max_Impact_Score = max(Impact_Score, na.rm = TRUE)
  )
```

All variables show that no NA values are present, and they were all cleaned in previous questions. Therefore, no further cleaning is needed.

#### Categorical Factors Investigation

#### 7b(ii) 1) Fate Investigation

```{r}
# Calculating impact_score by Fate
score_by_fate <- wastes_joined %>%
  group_by(Fate) %>%
  summarise(
    Mean_Score = mean(Impact_Score, na.rm = TRUE),
    Median_Score = median(Impact_Score, na.rm = TRUE),
    SD_Score = sd(Impact_Score, na.rm = TRUE),
    N = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(Mean_Score))

score_by_fate
```

```{r}
# Plotting a boxplot to visualise the distribution

ggplot(
  wastes_joined,
  aes(
    x = Fate,
    y = Impact_Score
  )
) +
  geom_boxplot() +
  coord_flip() +
  labs(
    title = "Environmental impact score by waste fate",
    x = "Waste fate",
    y = "Environmental impact score"
  ) +
  theme_minimal()
```

Waste fate shows a clear association with environmental impact score. Recycling has the highest mean score of 6.18, compared with 4.75 for disposal, suggesting that recycling cases tend to receive more positive environmental assessments. The boxplot also shows substantially greater variation for recycling and disposal, whereas energy recovery, long-term storage and waste reuse are concentrated entirely around score 5. An odd observation is disposal having a median of 6 despite its lower mean of 4.75, indicating that there are larger number of lower-scoring disposal cases which pull the average downward. Overall, Fate appears to be an important explanatory factor, although the very small observations of long-term storage and waste reuse cases mean these groups should be interpreted cautiously as they are lacking in data.

#### 7b(ii). 2) Stream Investigation

```{r}
# Calculating impact_score by Stream
score_by_stream <- wastes_joined %>%
  group_by(Stream) %>%
  summarise(
    Mean_Score = mean(Impact_Score, na.rm = TRUE),
    Median_Score = median(Impact_Score, na.rm = TRUE),
    N = n(),
    .groups = "drop"
  )

score_by_stream
```

```{r}
# Plotting a boxplot to visualise the distribution
ggplot(wastes_joined,
       aes(x = Stream, y = Impact_Score)) +
  geom_boxplot() +
  coord_flip() +
  labs(
    title = "Environmental Impact Scores by Waste Stream",
    x = "Waste Stream",
    y = "Environmental Impact Score"
  ) +
  theme_minimal()
```

Environmental impact scores differ only moderately across waste streams. MSW has the highest mean score at 5.61, followed by C&D at 5.43 and C&I at 5.22, while all three streams have a median score of 5. The boxplot indicates that MSW contains a greater proportion of higher impact scores, whereas C&I displays a wider spread toward lower values. However, the relatively small differences in the means suggest that Stream alone is unlikely to explain a large proportion of the variation in environmental impact scores.

#### 7b(ii.) 3) Category Investigation

```{r}
# Calculating impact_score by Category
score_by_category <- wastes_joined %>%
  group_by(Category) %>%
  summarise(
    Mean_Score = mean(Impact_Score, na.rm = TRUE),
    Median_Score = median(Impact_Score, na.rm = TRUE),
    N = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(Mean_Score))

score_by_category
```

```{r}
# Plotting a boxplot to visualise the distribution
ggplot(wastes_joined,
       aes(x = reorder(Category, Impact_Score, FUN = median),
           y = Impact_Score)) +
  geom_boxplot() +
  coord_flip() +
  labs(
    title = "Environmental Impact Scores by Waste Category",
    x = "Waste Category",
    y = "Environmental Impact Score"
  ) +
  theme_minimal()
```

Waste Category shows quite some great differences in environmental impact scores. Metals has the highest mean score of 6.70 and a median of 8, indicating that metal waste processes are concentrated toward more positive environmental effects. Glass, plastics and building and demolition materials also have relatively high average scores above 6. In contrast, Organics records the lowest mean score of 4.85. The boxplots also suggest that the material being processed is meaningfully associated with environmental impact. However, the group sizes are highly unequal, particularly because hazardous waste accounts for 28046 observations and Organics with 4213 observations, while other categories' obervations are in the hundreds. Therefore, category comparisons should be interpreted alongside a combined regression rather than from means alone.

#### 7b(ii.) 4) Core_Non-core Investigation

```{r}
# Calculating impact_score by Core_Non-core
score_by_core <- wastes_joined %>%
  group_by(`Core_Non-core`) %>%
  summarise(
    Mean_Score = mean(Impact_Score, na.rm = TRUE),
    Median_Score = median(Impact_Score, na.rm = TRUE),
    N = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(Mean_Score))

score_by_core
```

Non-core waste has a higher mean environmental impact score of 6.03 compared with 5.34 for core waste. However, this comparison should be treated cautiously because the groups are extremely imbalanced: only 125 observations are classified as non-core waste compared with 36563 core-waste observations. Therefore, the raw difference in means does not by itself provide strong evidence that core classification independently influences environmental impact.

#### 7b(ii.) 5) State Investigation

```{r}
# Calculating impact_score by State
score_by_state <- wastes_joined %>%
  group_by(State) %>%
  summarise(
    Mean_Score = mean(Impact_Score, na.rm = TRUE),
    Median_Score = median(Impact_Score, na.rm = TRUE),
    N = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(Mean_Score))

score_by_state
```

Geographic differences in environmental impact scores are relatively small. Vic records the highest mean score at 5.53, whereas ACT records the lowest at 5.17. All states have the same median score of 5. The narrow range of approximately 0.36 points suggests that State is associated with some variation in scores, but the differences are smaller than those observed for Fate and Category.

#### Numerical Factors Investigation

#### 7b(iii.) 1) Tonnes Investigation

As Tonnes have a distribution of zero to millions of tonnes as inspected previously, it is very likely to be highly skewed if plotted as it is. Therefore, log1p() will be used to transform the waste tonnes into logarithmic to show its relationship.

```{r}
# Calculating the correlation of Tonnes and Impact Score
tonnes_cor <- cor(
  log1p(wastes_joined$Tonnes),
  wastes_joined$Impact_Score,
  use = "complete.obs"
)

tonnes_cor
```

```{r}
# Plotting a scatter plot and linear regression to show the relationship
ggplot(wastes_joined,
       aes(x = log1p(Tonnes),
           y = Impact_Score)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Waste Quantity and Environmental Impact Score",
    x = "log(1 + Tonnes)",
    y = "Environmental Impact Score"
  ) +
  theme_minimal()
```

The correlation between log-transformed waste tonnes and environmental impact score is r=0.078, indicating a very weak positive linear relationship. The fitted regression line slopes slightly upward, but the effect is minimal. The scatterplot also shows strong horizontal clustering around impact scores 3, 5 and 8, reflecting the discrete distribution of the response variable rather than a clear relationship with waste quantity. Therefore, the scale of the waste process, measured by Tonnes, provides little explanatory power when considered by itself.

#### 7b(iii.) 2) Year Investigation

```{r}
# Calculating impact_score by Year
score_by_year <- wastes_joined %>%
  group_by(Year) %>%
  summarise(
    Mean_Impact_Score = mean(Impact_Score, na.rm = TRUE),
    .groups = "drop"
  )

score_by_year
```

```{r}
# Plotting a line graph to show the relationship
ggplot(score_by_year,
       aes(x = Year, y = Mean_Impact_Score)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Mean Environmental Impact Score Over Time",
    x = "Year",
    y = "Mean Environmental Impact Score"
  ) +
  theme_minimal()
```

The mean environmental impact scores show a gradual upward trend over the study period. The average increases from 5.24 in 2006 to 5.46 in 2020, reaching a peak of 5.49 in 2019. This indicates that environmental impact assessments became slightly more positive over time. However, the total increase is only about 0.22 points across the full period, so the magnitude of the change is small. Year may therefore have a statistically detectable association with impact score without necessarily representing a large practical effect.

#### 7b(iii.) 3) Economic Growth Investigation

Economic_Growth belongs to a State-Year combination and is repeated across many waste rows, therefore the investigation should be grouped by State-Year.

```{r}
# Calculating impact_score by Economic Growth
score_by_economic <- wastes_joined %>%
  group_by(State, Year, Economic_Growth) %>%
  summarise(
    Mean_Impact_Score = mean(Impact_Score, na.rm = TRUE),
    .groups = "drop"
  )

score_by_economic
```

```{r}
# Plotting a scatterplot and linear regression to show the relationship
ggplot(score_by_economic,
       aes(x = Economic_Growth,
           y = Mean_Impact_Score)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Economic Growth and Environmental Impact Score",
    x = "Economic Growth (%)",
    y = "Mean Environmental Impact Score"
  ) +
  theme_minimal()
```

```{r}
# Calculating the correlation of Economic Growth and Impact Score
economic_cor <- cor(
  score_by_economic$Economic_Growth,
  score_by_economic$Mean_Impact_Score,
  use = "complete.obs"
)

economic_cor
```

Economic growth has a weak negative correlation with mean environmental impact score at the State-Year level (r=-0.177). The scatterplot shows a slight downward fitted trend, suggesting that State-Year observations with higher economic growth tend to have lower environmental impact scores. However, the relationship is weak and the observations remain widely dispersed around the fitted line. Economic growth therefore appears to have limited explanatory power compared with more direct waste-process characteristics such as Fate and Category.

### 7c. Combined Regression

Multiple regression was used to examine the relationship between environmental impact score and several factors simultaneously. While the earlier descriptive analyses show how each factor is associated with the score individually, these relationships may not be the only explanation. Multiple regression helps isolate the association of each factor while controlling for the others, providing a clearer view of which variables remain important predictors of environmental impact score.

```{r}
# Forming the combined regression model for Impact Score
impact_model <- lm(
  Impact_Score ~ log1p(Tonnes) +
    Fate +
    Stream +
    Category +
    `Core_Non-core` +
    State +
    Year,
  data = wastes_joined
)

summary(impact_model)
```

From the impact_model, we can see that R^2^ = 0.2316 and Adjusted R^2^ = 0.231. Overall, the model shows p \< 2.2 x 10^-16^.

The multiple linear regression is statistically significant overall (p\<0.001), indicating that the included factors jointly provide information about environmental impact scores. However, the adjusted R^2^ of 0.231 indicates that the model explains only approximately 23.1% of the observed variation in impact scores. Therefore, while the available waste characteristics are associated with environmental impact, a large proportion of the variation remains unexplained by the variables in the dataset.

**Fate**

The referenced category is Disposal.

Fate remains one of the strongest factors after controlling for the other variables. Recycling is associated with an environmental impact score approximately 1.36 points higher than disposal on average, while energy recovery is associated with an increase of approximately 0.39 points. Both relationships are statistically significant. In contrast, waste reuse is not statistically different from disposal (p=0.640), although this result should be interpreted cautiously because waste reuse contains very few observations. These results further supports the earlier finding that the ultimate destination of waste is strongly associated with its environmental impact assessment.

**Stream**

The referenced category is C&D.

After controlling for other variables, both C&I and MSW have lower estimated environmental impact scores than the C&D reference stream. C&I is approximately 0.31 points lower and MSW approximately 0.12 points lower. This differs from the raw descriptive results, where MSW had the highest mean score. The change suggests that the raw stream differences were partly explained by differences in the types of materials and treatment methods occurring within each stream.

**Category**

The referenced category is Ash.

Waste Category also remains associated with environmental impact score after controlling for the other factors. Relative to the Ash reference category, Organics has the largest negative coefficient at approximately −1.32 points, followed by unclassified materials and textiles, leather and rubber. This supports the previous analysis showing that Organics has one of the lowest average impact scores. The category results suggest that the type of material being processed contributes meaningfully to environmental impact assessments even after treatment pathway, stream, state and year are taken into account.

**Tonnes**

The log-transformed Tonnes has a positive and statistically significant coefficient of approximately 0.035. However, the magnitude is very small, which is consistent with the earlier correlation of only 0.078. Therefore, waste quantity may have a detectable relationship with environmental impact score because of the large sample size, but its practical explanatory effect appears limited.

**State**

The referenced category is ACT.

Some geographic differences remain after controlling for waste characteristics. Relative to ACT, NSW, Victoria and WA have estimated scores approximately 0.25, 0.23 and 0.15 points higher respectively. Queensland does not differ significantly from ACT (p=0.375). Although several state effects are statistically significant, their magnitudes are relatively small, supporting the earlier conclusion that geographic location is a secondary factor compared with Fate and Category.

**Year**

Year has a positive and statistically significant coefficient of approximately 0.0126, meaning that environmental impact scores increase by around 0.013 points per year on average after controlling for the other included variables. Although the relationship is statistically significant, the annual effect is small, which is consistent with the gradual trend observed in the previous analysis.

**Core_Non-core**

The regression output shows NA for Core_Non-score and R mentions that '1 not defined because of singularities'. Core_Non-core could not be independently estimated in the multiple regression because the model identified a singularity, which indicates a perfect collinearity with another included predictor. Therefore, although Core/Non-core was useful for descriptive comparison, its independent effect cannot be separated from the other variables in this model.

### Q7 Answer and Justification

Overall, the analysis indicates that environmental impact scores are associated with several waste-process characteristics, but their explanatory strength differs. Fate and Category show the clearest relationships, where recycling has substantially higher scores than disposal after controlling for other factors, while Organics is associated with comparatively lower scores. Stream also contributes to score differences, although its raw relationship changes after other variables are controlled for, demonstrating the importance of considering multiple factors simultaneously. Waste quantity and Year are statistically significant but have relatively small effect sizes, while State shows slight geographic variation. Economic growth only has a weak negative relationship with mean impact score at the State-Year level. The regression model explains approximately 23.1% of the variation in environmental impact scores, indicating that the available variables provide useful but incomplete explanatory power. The findings therefore suggest that **how waste is managed** and **what material is being processed** are more strongly associated with environmental impact scores than waste quantity, geographic location or broader economic conditions. These relationships should be interpreted as associations rather than direct causal effects.

## Q8

**A manager is deciding where to invest in large-scale tyre processing capacity, and she wants
to understand the disposal size variation and disposal frequency for each state. Create a
relevant plot to support the manager's decision. Justify your choice, and explain what the plot
tells her. (5 marks)**

### 8a. Extracting Tyre Disposals

The Type and Tonnes have already been inspected and cleaned in previous questions, therefore no further cleaning is needed.

```{r}
# Filter out tyre disposal records
tyre_disposal <- wastes_joined %>%
  filter(
    Type == "Tyres (T140)",
    Fate == "Disposal"
  )

# Check for 0 tonne records
tyre_disposal %>%
  summarise(
    Total_Records = n(),
    Zero_Tonnes = sum(Tonnes == 0),
    Positive_Tonnes = sum(Tonnes > 0)
  )
```

To extract the actual tyre disposals that have occured, records with 0 tonnes need to be excluded from the analysis as they do not count as an actual tyre disposal.

```{r}
# Filtering out valid tyre disposal records
tyre_disposal <- tyre_disposal %>%
  filter(Tonnes > 0)
```

### 8b. Deciding Tonnes ranges

```{r}
# Inspecting Tonnes distribution
summary(tyre_disposal$Tonnes)
```

As observed from the statistical information of the valid tyre disposal tonne values, we can see that the value ranges from 86 to 70411, while the median is 6211. From this, we can conclude that half of the values are in the lower range, less than 6211. Therefore, a valid Tonne range would be to allocate a smaller range for the smaller values, and gradually growing the range for larger values.

```{r}
# Allocating each record into a Tonne range
tyre_disposal <- tyre_disposal %>%
  mutate(
    Tonnes_range = case_when(
      Tonnes <= 1000 ~ "0-1,000",
      Tonnes <= 5000 ~ "1,001-5,000",
      Tonnes <= 10000 ~ "5,001-10,000",
      Tonnes <= 50000 ~ "10,001-50,000",
      Tonnes > 50000 ~ ">50,000"
    )
  )

# Inspecting the Tonne distribution across the ranges
tyre_disposal %>%
  count(Tonnes_range)
```

### 8c. Calculating disposal frequency

```{r}
# Calculating disposal frequency by State
tyre_frequency <- tyre_disposal %>%
  count(
    State,
    Tonnes_range,
    name = "Frequency"
  )

tyre_frequency
```

```{r}
# Calculating disposal summary for each State
tyre_state_summary <- tyre_disposal %>%
  group_by(State) %>%
  summarise(
    Disposal_Frequency = n(), # Total disposal frequency
    Total_Tonnes = sum(Tonnes), # Total tonnes disposed
    Mean_Tonnes = mean(Tonnes),
    Max_Tonnes = max(Tonnes),
    .groups = "drop"
  ) %>%
  arrange(desc(Total_Tonnes))

tyre_state_summary
```

### 8d. Visualising the Statistics

A stacked bar chart was selected because the manager needs to compare both tyre disposal frequency and disposal-size variation across states. The total height of each bar represents the number of actual disposal events in each state, while the stacked segments show how those events are distributed across different Tonnes_range categories. This allows states with frequent disposal and a greater concentration of large disposal events to be identified at the same time.

```{r}
# Visualising the disposal size of each State
ggplot(
  tyre_frequency,
  aes(
    x = reorder(State, Frequency, FUN = sum),
    y = Frequency,
    fill = Tonnes_range
  )
) +
  geom_col(position = "stack") +
  coord_flip() +
  labs(
    title = "Tyre Disposal Frequency and Size Variation by State",
    x = "State",
    y = "Number of Disposal Events",
    fill = "Disposal Size (Tonnes)"
  ) +
  theme_minimal()
```

The plot shows that tyre disposal frequency is identical across all states, with 19 positive-tonnage disposal events each. Therefore, frequency does not differentiate the states for investment purposes. Instead, the important variation is the size of the disposal events, shown by the distribution of the Tonne range segments. New South Wales, Queensland, Victoria and Western Australia have substantially larger disposal quantities, whereas South Australia, Tasmania, Northern Territory and ACT generally have much smaller disposal sizes. This indicates that the manager should focus more on the size distribution of disposal events rather than their frequency when considering where large-scale processing capacity may be required.

### Q8 Answer and Justification

Tyre disposal frequency is constant across all eight states, with 19 positive-tonnage disposal records per state. Therefore, frequency alone does not provide an insight for choosing where to invest in processing capacity. The key difference is instead the size of the disposal events. NSW records the largest total tyre disposal volume at approximately 909422 tonnes, with an average of 47864 tonnes per event and a maximum of 70411 tonnes. Qld follows with approximately 783599 tonnes in total and an average of 41242 tonnes, while Vic and WA also record relatively high average disposal sizes of approximately 33340 and 31474 tonnes, respectively. In contrast, SA, Tas, NT and ACT have substantially smaller average disposal quantities.

Since disposal frequency is constant across states, the stronger evidence for large-scale processing demand comes from disposal size. **NSW shows the greatest potential demand for large-scale tyre processing capacity**, as it has the highest total disposal volume, highest average disposal size and largest single disposal event. Queensland is the next strongest candidate, followed by Victoria and Western Australia. The lower average and maximum disposal quantities in South Australia, Tasmania, Northern Territory and Australia Capital Territory suggest less need for very large processing capacity based on this dataset.

Additionally, across all states, 71 of the 152 disposal events fall above 10,000 tonnes (42 in the 10,001–50,000 range and 29 above 50,000 tonnes). This indicates that large disposal events are quite common overall, which suggests the demand for large-scale processing capacity.
