---
title: "34474056_A2"
author: "Chan Long Hei"
date: "2026-09-6"
output: html_document
---

# FIT5145 Assignment 2

## Loading Libraries and Datasets

```{r}
library(tidyverse)
library(stringr)
library(scales)

# Read files
wastes <- read_csv("Wastes.csv")
year_state <- read_csv("Year_State_ID.csv")
```

## Initial Inspection of Datasets

```{r}
# Roughly look at the datasets
glimpse(wastes)
glimpse(year_state)

# Looking at summary to see if anything is out of line
summary(wastes)

# Looking for NA entries for each columns
colSums(is.na(wastes))
```

## **Q1**

**How many unique “Category” values are there in the data file (6 marks)?**

### 1a. Inspecting raw data for Category values

```{r}
# Inspect all Category values and their frequencies
wastes %>%
  count(Category, sort = TRUE)
```

### 1b. Checking other Types to reassign some suspicious Category values

```{r}
# Extracting suspicious Category
suspicious_categories <- c(
  "organics",
  "Plastic",
  "Building & demolition materials",
  "HW",
  "Toxic wastes",
  "Aluminium",
  "Metals & Organics",
  "Recycling",
  "Space debris"
)

# Inspecting suspicious Category values
wastes %>%
  filter(
    Category %in% suspicious_categories |
      is.na(Category)
  ) %>%
  select(
    Case_ID,
    Category,
    Type,
    Stream,
    Fate,
    Tonnes
  )
```

```{r}
# Find the Types associated with the suspicious Category values
suspicious_types <- wastes %>%
  filter(Category %in% suspicious_categories) %>%
  distinct(Category, Type) %>%
  arrange(Category, Type)

suspicious_types
```

```{r}
# Verifying the common Category associated with these types
wastes %>%
  semi_join(
    suspicious_types,
    by = "Type"
  ) %>%
  count(Type, Category) %>%
  arrange(Type, desc(n))
```

```{r}
# Find Types where Category is missing
missing_category_types <- wastes %>%
  filter(is.na(Category)) %>%
  distinct(Type)

missing_category_types
```

```{r}
# Check where those Types normally belong
wastes %>%
  semi_join(
    missing_category_types,
    by = "Type"
  ) %>%
  filter(!is.na(Category)) %>%
  count(Type, Category, sort = TRUE)
```

### 1c. Cleaning the Category values

```{r}
wastes_clean <- wastes %>%
  mutate(
    Category = case_when(
      
      # Naming inconsistencies
      Category == "organics" ~ "Organics",
      Category == "Plastic" ~ "Plastics",
      Category == "Building & demolition materials" ~
        "Building and demolition materials",
      
      # Wrongly classified values verified using Type
      Category == "Aluminium" ~ "Metals",
      Category == "Metals & Organics" ~ "Organics",
      Category == "Recycling" ~ "Plastics",
      Category == "HW" ~ "Hazardous wastes",
      Category == "Toxic wastes" ~ "Hazardous wastes",
      
      # Missing categories that can be inferred from their Type
      is.na(Category) &
        Type == "Other soil/sludges (other N)" ~
        "Hazardous wastes",
      
      is.na(Category) &
        Type == "Paints, resins, inks, organic sludges (F)" ~
        "Hazardous wastes",
      
      # Keep all otherwise valid values unchanged
      TRUE ~ Category
    )
  )

# Removing the entry where Category is Space Debris
wastes_clean <- wastes_clean %>%
  filter(
    !(Category == "Space debris" &
        is.na(Case_ID) &
        is.na(Year_State_ID))
  )
```

### 1d. Calculate the number of unique Categories

```{r}
# Showing all unique Categories
wastes_clean %>%
  distinct(Category) %>%
  arrange(Category)
```

```{r}
# Showing number of unique Categories
wastes_clean %>%
  summarise(
    unique_categories =
      n_distinct(Category, na.rm = TRUE)
  )
```

### Q1 Answer and Justification

Initial inspection found 21 distinct values in the raw Category column including NA. However, several values represented inconsistent naming or apparent data-entry errors rather than genuinely different waste categories. For example, organics and Organics, Plastic and Plastics, and the two forms of building and demolition materials referred to the same categories. Other suspicious values were checked against their corresponding Type. Aluminium was reassigned to Metals, which is supported both by other observations and by the specification, which describes Aluminium as a Type within the Metals category. Similarly, Recycling was reassigned to Plastics because its Type was PVC and the specification defines recycling as a Fate rather than a Category. Missing Category values were assigned to Hazardous wastes where the same Types were consistently classified as hazardous waste elsewhere in the data. One incomplete Space debris record was excluded because it had no case ID, year/state ID, stream, fate, tonnage or description and therefore could not be meaningfully used in the analysis. After these corrections, the dataset contains **11 unique waste Categories**.

## Q2

**How many negative feedback comments are in the 'Description' column with an
environmental impact score of 2 or 3 (4 marks)?**

From initial inspection of the raw dataset and data cleaning in Q1, we know that there is 1 NA entry in the Description column in the raw dataset, which is associated with the entry with Category "Space Debris". It was already removed back in Q1, therefore we can conclude that there is no more NA entry in the Description column.

### 2a. Inspection of Description structure

```{r}
# Extracting 10 rows to inspect the entry structure
wastes_clean %>%
  select(Description) %>%
  slice_head(n = 10)
```

### 2b. Extracting Environmental Impact Score

```{r}
q2_data <- wastes_clean %>%
  mutate(
    Impact_Score = as.numeric(
      str_extract(
        Description,
        "(?<=Environmental Impact Score: )[0-9]+"
      )
    )
  )

q2_data
```

### 2c. Extracting negative feedback comments

```{r}
# Showing number of rows in each of the impact score of 2,3
q2_data %>%
  filter(Impact_Score %in% c(2, 3)) %>%
  count(Impact_Score)
```

```{r}
# Calculating total number of negative feedback comments
q2_data %>%
  filter(Impact_Score %in% c(2, 3)) %>%
  summarise(
    negative_feedback_comments = n()
  )
```

### Q2 Answer and Justification

Since the only missing Description record had already been removed during the earlier cleaning stage in Q1, no additional missing-value treatment was required. The environmental impact score was extracted from the text in Description, then records with scores of 2 or 3 were filtered and counted. There are a total of **6312 negative feedback comments** in the Description column with an environmental impact score of 2 or 3.
