---
title: "34474056_A2"
author: "Chan Long Hei"
date: "2026-09-6"
output: html_document
---

# FIT5145 Assignment 2

## Loading Libraries and Datasets

```{r}
library(tidyverse)
library(stringr)
library(scales)

# Read files
wastes <- read_csv("Wastes.csv")
year_state <- read_csv("Year_State_ID.csv")
```

## Initial Inspection of Datasets

```{r}
# Roughly look at the datasets
glimpse(wastes)
glimpse(year_state)
```

```{r}
# Looking at summary to see if anything is out of line
summary(wastes)
summary(year_state)
```

```{r}
# Looking for NA entries for each column in wastes
colSums(is.na(wastes))

```

```{r}
# Looking for NA entries for each column in year_state
colSums(is.na(year_state))
```

## **Q1**

**How many unique “Category” values are there in the data file (6 marks)?**

### 1a. Inspecting raw data for Category values

```{r}
# Inspect all Category values and their frequencies
wastes %>%
  count(Category, sort = TRUE)
```

### 1b. Checking other Types to reassign some suspicious Category values

```{r}
# Extracting suspicious Category
suspicious_categories <- c(
  "organics",
  "Plastic",
  "Building & demolition materials",
  "HW",
  "Toxic wastes",
  "Aluminium",
  "Metals & Organics",
  "Recycling",
  "Space debris"
)

# Inspecting suspicious Category values
wastes %>%
  filter(
    Category %in% suspicious_categories |
      is.na(Category)
  ) %>%
  select(
    Case_ID,
    Category,
    Type,
    Stream,
    Fate,
    Tonnes
  )
```

```{r}
# Find the Types associated with the suspicious Category values
suspicious_types <- wastes %>%
  filter(Category %in% suspicious_categories) %>%
  distinct(Category, Type) %>%
  arrange(Category, Type)

suspicious_types
```

```{r}
# Verifying the common Category associated with these types
wastes %>%
  semi_join(
    suspicious_types,
    by = "Type"
  ) %>%
  count(Type, Category) %>%
  arrange(Type, desc(n))
```

```{r}
# Find Types where Category is missing
missing_category_types <- wastes %>%
  filter(is.na(Category)) %>%
  distinct(Type)

missing_category_types
```

```{r}
# Check where those Types normally belong
wastes %>%
  semi_join(
    missing_category_types,
    by = "Type"
  ) %>%
  filter(!is.na(Category)) %>%
  count(Type, Category, sort = TRUE)
```

### 1c. Cleaning the Category values

```{r}
wastes_clean <- wastes %>%
  mutate(
    Category = case_when(
      
      # Naming inconsistencies
      Category == "organics" ~ "Organics",
      Category == "Plastic" ~ "Plastics",
      Category == "Building & demolition materials" ~
        "Building and demolition materials",
      
      # Wrongly classified values verified using Type
      Category == "Aluminium" ~ "Metals",
      Category == "Metals & Organics" ~ "Organics",
      Category == "Recycling" ~ "Plastics",
      Category == "HW" ~ "Hazardous wastes",
      Category == "Toxic wastes" ~ "Hazardous wastes",
      
      # Missing categories that can be inferred from their Type
      is.na(Category) &
        Type == "Other soil/sludges (other N)" ~
        "Hazardous wastes",
      
      is.na(Category) &
        Type == "Paints, resins, inks, organic sludges (F)" ~
        "Hazardous wastes",
      
      # Keep all otherwise valid values unchanged
      TRUE ~ Category
    )
  )

# Removing the entry where Category is Space Debris
wastes_clean <- wastes_clean %>%
  filter(
    !(Category == "Space debris" &
        is.na(Case_ID) &
        is.na(Year_State_ID))
  )
```

### 1d. Calculate the number of unique Categories

```{r}
# Showing all unique Categories
wastes_clean %>%
  distinct(Category) %>%
  arrange(Category)
```

```{r}
# Showing number of unique Categories
wastes_clean %>%
  summarise(
    unique_categories =
      n_distinct(Category, na.rm = TRUE)
  )
```

### Q1 Answer and Justification

The initial inspection shows that there are 21 distinct values in the raw Category column including NA. However, several values represented inconsistent naming or apparent data-entry errors rather than genuinely different waste categories. For example, organics and Organics, Plastic and Plastics, and the two forms of building and demolition materials referred to the same categories. Other suspicious values were checked against their corresponding Type. I have decided to impute them from other types instead of removing them from the dataset. Aluminium was reassigned to Metals, which is supported both by other observations and by the specification, which describes Aluminium as a Type within the Metals category. Similarly, Recycling was reassigned to Plastics because its Type was PVC and the specification defines recycling as a Fate rather than a Category. Missing Category values were assigned to Hazardous wastes where the same Types were consistently classified as hazardous waste elsewhere in the data. One incomplete Space debris record was excluded because it has more than 2 missing variables, with no case ID, year/state ID, stream, fate, tonnage or description and therefore could not be meaningfully used in the analysis. After these corrections, the dataset contains **11 unique waste Categories**.

## Q2

**How many negative feedback comments are in the 'Description' column with an environmental impact score of 2 or 3 (4 marks)?**

From initial inspection of the raw dataset and data cleaning in Q1, it was found that there is 1 NA entry in the Description column in the raw dataset, which is associated with the entry with Category "Space Debris". It was already removed back in Q1, therefore we can conclude that there is no more NA entry in the Description column.

### 2a. Inspection of Description structure

```{r}
# Extracting 10 rows to inspect the Description structure
wastes_clean %>%
  select(Description) %>%
  slice_head(n = 10)
```

### 2b. Extracting Environmental Impact Score

```{r}
q2_data <- wastes_clean %>%
  mutate(
    Impact_Score = as.numeric(
      str_extract(
        Description,
        "(?<=Environmental Impact Score: )[0-9]+"
      )
    )
  )

q2_data
```

### 2c. Extracting negative feedback comments

```{r}
# Showing number of rows in each of the impact score of 2,3
q2_data %>%
  filter(Impact_Score %in% c(2, 3)) %>%
  count(Impact_Score)
```

```{r}
# Calculating total number of negative feedback comments
q2_data %>%
  filter(Impact_Score %in% c(2, 3)) %>%
  summarise(
    negative_feedback_comments = n()
  )
```

### Q2 Answer and Justification

Since the only missing Description record had already been removed during the earlier cleaning stage in Q1, no additional missing-value treatment was required. The environmental impact score was extracted from the text in Description, then records with scores of 2 or 3 were filtered and counted. There are a total of **6312 negative feedback comments** in the Description column with an environmental impact score of 2 or 3.

## Q3

**For each Category value, write code to calculate the fractions of waste tonnes of different
waste sources, and then draw a chart to visualise the fraction numbers specific to the Category
value (4 marks).**

### 3a. Data inspection and cleaning

From the initial inspection of the dataset, it was found that there were 15 missing Tonnes values. It was also found that there are negative Tonnes values, which should be further investigated.

```{r}
# Calculating total missing Tonnes values and their contribution to the entire dataset
wastes_clean %>%
  summarise(
    missing_tonnes = sum(is.na(Tonnes)),
    missing_pct = mean(is.na(Tonnes)) * 100
  )
```

```{r}
# Inspecting negative Tonnes values
wastes_clean %>%
  filter(Tonnes < 0) %>%
  select(
    Case_ID,
    Category,
    Type,
    Stream,
    Fate,
    Tonnes,
    Description
  )
```

From the initial inspection of the dataset, it was found that there was 1 missing Stream value, which was removed back in Q1. Therefore, no further NA checking has to be done.

```{r}
# Inspecting the Stream variable
wastes_clean %>%
  count(Stream)
```

```{r}
# Removing the entries with missing Tonnes, Replacing the odd negative Tonnes value and cleaning up the inconsistent Stream entries
wastes_clean <- wastes_clean %>%
  mutate(
    Stream = case_when(
      Stream == "Municipal solid waste" ~ "MSW",
      Stream == "Construction and demolition waste" ~ "C&D",
      TRUE ~ Stream
    ),
    Tonnes = case_when(
      Case_ID == 34666 ~ 141663.39,
      Tonnes < 0 & abs(Tonnes) < 0.001 ~ 0,
      TRUE ~ Tonnes
    )
  ) %>%
  filter(!is.na(Tonnes))
```

### 3b. Calculating Waste Tonnes Fractions

```{r}
# Calculate total tonnes and fractions by Category and Stream
q3_summary <- wastes_clean %>%
  group_by(Category, Stream) %>%
  summarise(
    Total_Tonnes = sum(Tonnes),
    .groups = "drop"
  ) %>%
  group_by(Category) %>%
  mutate(
    Fraction = Total_Tonnes / sum(Total_Tonnes)
  ) %>%
  ungroup() %>%
  mutate(
    Category = factor(
      Category,
      levels = sort(unique(Category))
    )
  )

# Display the calculated fractions
q3_summary
```

```{r}
# Check fractions sum to 1 within every Category
q3_summary %>%
  group_by(Category) %>%
  summarise(
    Total_Fraction = sum(Fraction)
  )
```

### 3c. Visualising the fractions 

```{r}
# Visualise the fractions
ggplot(
  q3_summary,
  aes(
    x = Category,
    y = Fraction,
    fill = Stream
  )
) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(
    labels = scales::percent_format(),
    limits = c(0, 1)
  ) +
  labs(
    title = "Fraction of Waste Tonnes by Source for Each Category",
    x = "Waste Category",
    y = "Fraction of Total Waste Tonnes",
    fill = "Waste Source"
  ) +
  theme_minimal()
```

### Q3 Answer and Justification

Before calculating the fractions, the Tonnes variable was checked for missing values. Fifteen observations had missing tonnage values, representing approximately 0.04% of the dataset. Since only a very small fraction of cases was affected, these observations were removed rather than imputed.

Inspection of negative Tonnes values identified one odd case: Case 34666 contained a value of -20,000,000 tonnes, while its Description stated that 141,663.39 tonnes were processed. Since the description provided a clear corresponding quantity, the suspicious value was corrected to 141,663.39 rather than removing the observation. The remaining negative values were extremely small and close to 0, with their descriptions reporting -0.00 tonnes. These were treated as numerical precision errors and replaced with 0 because a negative waste quantity is not meaningful.

The Stream variable was standardised before analysis because the dataset contained duplicate labels for the same waste source, such as Municipal solid waste and MSW, and Construction and demolition waste and C&D. Since the specification defines only three stream categories, MSW, C&I, and C&D, these duplicate labels were recoded into the standard forms. This prevents the same waste source from being treated as separate groups when calculating fractions.

Total tonnes were then calculated for each combination of Category and Stream, and each stream total was divided by the total tonnes within its Category to obtain the fraction. A 100% stacked bar chart was selected because both Category and Stream are categorical variables and the values being visualised are proportions that sum to 100% within each Category. This allows the composition of waste sources to be compared directly across waste categories.

## Q4

**Please add the 'Year' and 'State' values from Year_State_ID.csv to Wastes.csv, compute the
total waste tonnes for each year and state, and store the result in a dataset named 'temp'. Then,
use a single R function to compute and display the statistical information (i.e., Min, Max, and
Mean) of the total waste tonnes for each state in ‘temp’ (Note: You may use multiple
functions/commands to prepare the pre-processed data table, but when you compute and
display the statistical information, you need to use a single R function.) (4 marks).**

### 4a. Inspecting year_state dataset

From the initial inspection of the dataset, it was found that there were no NA values for all columns in year_state.

```{r}
# Inspect all State values and their frequencies
year_state %>%
  count(State, sort = TRUE)
```

No duplicate states with inconsistent format are present.

```{r}
# Determining if there are duplicate ID values
n_distinct(year_state$ID)
nrow(year_state)
```

All year_state IDs are distinct.

```{r}
# Inspect all Year values and their frequencies
year_state %>%
  count(Year, sort = TRUE)
```

All Year values are formatted consistently.

```{r}
# Check if all Year_State_ID from wastes match to an existing ID in year_state
wastes_clean %>%
  # anti_join is used to identify mismatches found for the foreign key (wastes$Year_State_ID)
  anti_join(
    year_state,
    by = c("Year_State_ID" = "ID")
  )
```

All Year_State_ID in wastes has an existing match to an ID in year_state.

### 4b. Joining the datasets

```{r}
# Joining the datasets
wastes_joined <- wastes_clean %>%
  left_join(
    year_state,
    by = c("Year_State_ID" = "ID")
  )
```

```{r}
# Roughly look at the joined dataset
glimpse(wastes_joined)
```

```{r}
# Check if any rows were excluded during the join
nrow(wastes_clean)
nrow(wastes_joined)
```

The length of both the datasets matches, meaning that all entries are sucessfully matched with its respective year and state.

### 4c. Creating temp and displaying the statistical information

Tonnes values were already cleaned back in Q3, therefore no additional cleaning is needed before computing total waste tonnes.

```{r}
# Computing total waste tonnes for each year and state
temp <- wastes_joined %>%
  group_by(Year, State) %>%
  summarise(
    Total_Tonnes = sum(Tonnes),
    .groups = ("drop")
  )

temp
```

```{r}
# Reshaping temp so that each State is a column
temp <- temp %>%
  select(Year, State, Total_Tonnes) %>%
  pivot_wider(
    names_from = State,
    values_from = Total_Tonnes
  )

temp
```

```{r}
# Showing the statistical information of temp with a single function
summary(temp)
```

### Q4 Answer and Justification

The Year_State_ID dataset was first inspected to determine if any cleaning is needed, in which no additional cleaning was required. Wastes.csv and Year_State_ID.csv were then combined using a left join, matching Year_State_ID in the waste data with ID in the year-state data. A left join was used because Wastes.csv is the main dataset and the objective is to add the corresponding Year and State values from Year_State_ID.csv while retaining all waste observations. Year_State_ID in Wastes.csv acts as the foreign key, while ID in Year_State_ID.csv is the primary key. The joined data were grouped by Year and State to calculate total waste tonnes and stored as temp. Finally, summary() was used as the single R function to display the minimum, mean, and maximum total waste tonnes for each state.
